### 2025-06-08 manuscript/main.tex
cps, probProg

The thesis focuses on the above mentioned classes, as these are the ones applicable in the implementation part of the language service?NAME?. The last two classifications
\textbf{Type-1} and \textbf{Type-0} grammars, can create \textit{context-sensitive} and \textit{recursively enumerable} languages respectively, and are the most expressive
of all with Type-0 grammars placing absolutely no constraints on the production, making them only acceptable by Turing-Machine.

@@

Domain-Specific languages are programming or specification languages which are tailored to specific application domain, providing constructs and abstractions
that capture the domain concepts. \cite{whenHowDSL} In stark contrast to general-purpose languages (GLPs) that aim for broad applicability, a DSL waives generality in favor
of higher expressive power in its domain of interest. By offering notation closely related to a domains phrasing, a DSL allows solutions to be formulated at the same level
of abstraction as the domain itself, mapping language primitives onto the domain concepts. Developers, therefore, are able to create programs or specifications more
productively and securely.

The close relationship of DSL and domain furnishes multiple advantages over conventional GLPs: it reduces the amount of general programming knowledge needed, 
expanding the pool of developers from traditional programmers to also domain experts. Moreover, DSLs declutter code, rather than describing a subproblem of the domain
with multiple lines of code, fewer are needed because constructs inherently carry domain specific information, also contributing to broader comprehensibility for those familiar with the domain.
DSLs solutions are also easier to formally verify, as they are restricted to a limited context, than GLPs for which proving correctness is extremely hard.

@@

In domains where fine-grained control and explicit ordering matter, imperative DSLs provide maximal flexibility, because every aspect of execution is adjustable.
This level of control becomes crucial when timing or order of operations yield different outcomes, or when fine-tuned performance optimizations are required.
Furthermore, building an imperative DSL is often simpler than creating a declarative one: typically, an interpreters process each command in sequence, 
or a translator converts DSL instructions into a host general-purpose language.

@@

Imperative DSLs shine in scripting and macro contexts, where users sequence domain-specific commands to drive behavior, examples include Excel macros that
automate spreadsheet tasks or scripts within a game engine that dictate character actions and game logic. They are also well suited to testing frameworks
and scenario specifications, where user interfaces describe interactions by specifying a timely sequence of user inputs. In robotics and industrial automation, 
imperative DSLs allow practitioners to program custom action sequences—such as moving a robotic arm
to pick up an object, then placing it precisely—where each step must follow a strict order to ensure safety and accuracy. More broadly, any process-control 
domain that relies on precise sequencing and stateful interactions—be it a complex simulation workflow, an interactive operation pipeline, or automated machinery
—benefits from the clarity and control that an imperative DSL provides.

\chapter{Probabilistic Programming \& Cyber-Physical Systems}
\section{What Is Probabilistic Programming?}
\section{Inference and Sampling Methods}
\section{Cyber-Physical Systems: Definitions and Examples}
\section{Motivation for DSLs in P-P and CPS Contexts}

@@

\chapter{Full Grammar Specification}
% Full Miniprob grammar here.

>>>>

for choosing between replacing or recalibrating components—but measured data often reflect intertwined effects and closed-loop corrections. tit{context-sensitive} and \textit{recursively enumerable} languages respectively, and are the most expressive
of all with Type-0 grammars placing absolutely no constraints on the production, making them only acceptable by Turing-Machine.

@@

for choosing between replacing or recalibrating components—but measured data often reflect intertwined effects and closed-loop corrections.  to also domain experts. Moreover, DSLs declutter code, rather than describing a subproblem of the domain
with multiple lines of code, fewer are needed because constructs inherently carry domain specific information, also contributing to broader comprehensibility for those familiar with the domain.
DSLs solutions are also easier to formally verify, as they are restricted to a limited context, than GLPs for which proving correctness is extremely hard.

@@

In domains where fine-grained control and explicit ordering mat
for choosing between replacing or recalibrating components—but measured data often reflect intertwined effects and closed-loop corrections. reating a declarative one: typically, an interpreters process each command in sequence, 
or a translator converts DSL instructions into a host general-purpose language.

@@

Imperative DSLs shine in scripting and macro contexts, where users sequence domain-specific commands to drive behavior, examples include Excel macros that
automate spreadsheet tasks or scripts within a game engine that dictate character actions and game logic. They are also well suited to testing frameworks
and scenario specifications, where user interfaces describe interactions by specifying a timely sequence of user inputs. In robotics and industrial automation, 
imperative DSLs allow practitioners to program custom action sequences — such as moving a robotic arm
to pick up an object, then placing it precisely — where each step must follow a strict order to ensure safety and accuracy. More broadly, any process-control 
domain that relies on precise sequencing and stateful interactions — be it a complex simulation workflow, an interactive operation pipeline, or automated machinery
— benefits from the clarity and control that an impera
for choosing between replacing or recalibrating components—but measured data often reflect intertwined effects and closed-loop corrections. tive DSL provides.

\chapter{Probabilistic Programming \& Cyber-Physical Systems}
\section{What Is Probabilistic Programming?}

Probabilistic programming is a paradigm that aims to perform statistical analysis using tools yielded by computer science.\cite{introductionprobabilisticprogramming}
Unifying general-purpose programming and probabilistic modelling enables the presentation of statistical models as a program. Defined by underlying program code, the model is
embodied by the relations between variables and calculations. The established capabilities of programming languages are often augmented with 
probabilistic functionality, in ... creating probabilistic programming languages (PPL),\cite{probProgrammingPrinciples} that are closely associated with this paradigm. 
The enriched syntax enables the creation of more compact models and fosters conciser abstractions of the underlying probabilistic structure, usually providing constructs
that allow declaring random variables and conditioning of observed data, typical for probabilistic modelling. These constructs form the foundation for specifying core elements
of statistical models. \textit{Priors} are pre-known and usually drawn from distributions reflecting the initial belief over what value
\textit{latent variables}/model parameters might take on. \textit{Likelihoods} are also known beforehand representing how data is assumed to be generated given the 
latent variable for which the \textit{posterior} is to be inferred. Together, these components form the backbone of a probabilistic model and provide the information required
to compute the posterior distribution, which is grounded in \textbf{Bayesian theory}.\cite{introductionprobabilisticprogramming}
\\

\textit{Bayes’ Theorem} provides the foundational framework for probabilistic reasoning under uncertainty. It describes how to update prior beliefs about unknown 
quantities - known as latent variables - based on new evidence, resulting in a posterior distribution. Formally, it states that the posterior is proportional to the
product of the prior and the likelihood:
\[
P(\theta \mid x) \propto P(x \mid \theta) \cdot P(\theta)
\]
In probabilistic programming, this principle is applied through models that define prior distributions over latent variables and likelihood functions for 
observed data. The probabilistic programming language then applies Bayes' Theorem, typically via automated inference algorithms, to compute or approximate the posterior.
This makes it possible to express complex probabilistic models declaratively, while delegating the computation of posterior beliefs to the underlying inference engine.
\\
??INLUDE GRAPHICAL MODESL??

\section{Inference and Sampling Methods}

Inference is the process by which probabilistic models are "executed". Given a model and observed data, the task is to compute the posterior distributions of latent variables.
Because probabilistic models often include uncertainty and hidden variables, inference is the process that turns the model into concrete answers.
This process is abstracted and automated, but remains computationally nontrivial. Standard approaches to inference are generally grouped 
into three main categories- \textit{exact inference}, \textit{sampling-based methods}, and \textit{variational inference}.\cite{bishop2006prml, blei2017vi, goodman2014dippl}.

\subsection*{1. Exact Inference}

Exact inference computes the posterior distribution analytically or through complete enumeration. This is possible only in models with limited complexity where the 
entire state space can be traversed. Although conceptually straightforward, exact inference quickly becomes infeasible as model size increases, due to the combinatorial
explosion of possible states.

\subsection*{2. Sampling-Based Inference}

Sampling-based methods approximate the posteriIn probabilistic programming, t

Variational inference formulates inference as an optimization problem. Instead of sampling, it produces a family of approximate distributions and seeks the member that 
is closest to the true posterior. This involves defining an objective function - often the \textit{evidence lower bound (ELBO)} - and optimizing it using gradient-based methods.
Variational inference tends to be faster and more scalable than sampling, though it trades off exactness for speed by approximating the posterior rather than sampling
from it directly \cite{blei2017vi}.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.7\textwidth]{graphics/inferencePlot.png}
  \caption{Comparison of exact posterior to inference methods}
  \label{fig:inference_plot}
\end{figure}

\section{Cyber-Physical Systems: Definition and Examples}

Cyber-Physical System (CPS) is an umbrella term broadly referring to any systems which tightly integrate both computational logic (cyber) and physical components, allowing interplay
through designated interfaces. \cite{cpsModelDesc} The two parts of a CPS have to be seamlessly interconnected enabling efficient communication and consistent information
 relying?? \cite{cpsChallengesAndFuture}, as both sides heavily depend on one another.
The cyber-physical relationship can be generally describes by three roles \textit{controller/agent}, \textit{sensors} and \textit{activators},
which are connected by an underlying network.
Controllers, representing the 'cyber', are occupied with computation and relying fitting instructions based on te observational data provided by the physical sensors.
Sensor and Actuators embody the 'physical' and are the ones interacting with the environments in which the system resides. A sensors jobs is to gather intel from its
surroundings communicating it in order for the agent to make correct decisions and predictions and to monitor the system. Without sufficient data, the controller is left
blind and possibly  unable to perform as intended. While sensors passively interact with the environment by observing, actuators do so actively by altering the systems state
 — carrying out tasks instructed by the controller.
  ??Looping behaviour??

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.7\textwidth]{graphics/cpsGeneralArchitecture.png}
  \caption{General architecture of common CPS}
  \label{fig:cps_architecture}
\end{figure}

Controllers may range from simple reactive routines to advanced inference models grounded in probabilistic programming, or even intelligent agents capable of autonomous
decision-making. Actuators, in turn, translating these decisions into physical action can be represented by basic devices such as motors that move parts, 
valves that regulate flow, or relays that control circuits. They can also act as more complex assemblies — robotic joints, programmable hydraulic systems, or smart actuators 
that adjust based on local conditions. At the foundation, sensors continuously monitor the environment, capturing data through modalities such as motion, 
orientation, or spatial mapping. Together, these elements form  an integrated feedback system, enabling CPS to respond intelligently and flexibly to their physical context,
regardless of application domain.\\

CP-Systems are inherently multidisciplinary structures intersecting engineering domains (mechanical, electrical, civil, etc.) with computer science.
Because these domains employ differing models for ... , effectively combining them is non-trivial. Challenges arise during development and operation of CPSs for which
finding solutions is imperative. \cite{cpsChallengesAndFuture} First and foremost is the concern of \textbf{reliability and safety}. Occurrences of failures can have dire real-world
consequences with catastrophic end results. This means systems must handle unexpected conditions gracefully and robustly. But ensuring safety through extensive verification 
and validation is complicated by the combination of continuous physical dynamics and discrete decision logics. Another challenge is \textbf{real-time performance}, as computing
elements must keep pace with the physical processes, which usually require low-latency responsiveness including the network. Since CPSs are used in infrastructure and other
critical sites that may be targets of cyber attacks, guarantying a high standard of \textbf{Security} is essential. Additionally, sensors are not completely reliable — 
possibly introducing noise to measurements — and environmental conditions can vary drastically leading to \textbf{uncertainty}, which a CPS has to cope with to be resilient. \cite{cpsProbabilisticRobotics}

\section{Motivation for PPLs in CPS Contexts}
??enumerate the four paragrpahs into 2x2??

Probabilistic programming has multiple fitting applications within CPSs, especially where reasoning under uncertainty is essential. Generally, probabilistic programming
allows controllers to plan even in unpredictable environments and settings.

When measurements contain noise or are incorrect, PPL models can still be applied to extract useful information. Instead of treating the raw reading of a sensor as the “truth,”
the readings are considered samples for a \textit{sensor’s model}. Not only is it possible to reduce the noise in the measurements by calculating the posterior, 
but those noise characteristics are also used to flesh out the predictive model itself.
This enables PPLs to adaptively adjust noise models online when, for example, sensors deteriorate over time.
Additionally — with most CPSs employing multiple sensors — a PPL naturally supports \textit{sensor fusion}, \cite{cpsSensorFusion} combining data from heterogeneous sources
to produce a coherent estimate. By defining each sensor's observation as a conditional distribution, a probabilistic program
weights each reading according to its uncertainty: more reliable sensors contribute more heavily, while inconsistent or faulty readings are down-weighted or flagged. 
Inference then yields a joint posterior over the true state, allowing the system to reconcile disagreements between sensors, detect potential sensor failures by inferring 
latent bias variables, and maintain a quantified confidence.\\

Which leads us into robotics, a quintessential cornerstone of CPS. Instead of working on raw observational data and making assumptions in incomplete cases, robots can adopt 
a \textit{belief} based on an observational model \cite{cpsProbabilisticRobotics}. The observational model couples motion dynamics (how actions translate into state 
transitions) with existing sensor models, providing a full posterior distribution that the robot can use to imply different levels of confidence for its actions.
Furthermore, robots often operate in changing environments or with components that degrade over time. Within a PPL, unknown parameters, such as friction 
coefficients or sensor calibration offsets, can be treated as latent variables and perform Bayesian learning to infer them from incoming data. The same inference mechanism used for state 
estimation thus also adapts the model online, continuously refining both motion dynamics and sensor mappings.\\

A PPL expresses system dynamics probabilistically so that sampling yields a distribution of possible futures. \cite{cpsPredictiveControl} By enforcing “chance constraints” 
(e.g., “limit the probability of exceeding a safety threshold to 1\%”), inference or Monte Carlo sampling identifies control inputs that optimize expected performance while 
respecting risk. When the current state is uncertain, planning occurs in belief space — each action triggers a Bayesian update of the state distribution, and planners 
choose sequences that minimize expected cost given that uncertainty. Simultaneously, online Bayesian learning refines model parameters (like friction or drag) over time, 
tightening future predictions; as uncertainty shrinks, control can push closer to limits, whereas rising uncertainty leads to more conservative actions.\\

Recent research is looking at ways to utilize PPLs to gather diagnostics of a CPS. \cite{cpsPPLDiagnostics} They authors propose a way to infer potential error causes, that
might be overlooked otherwise, due to the . 
This of of great interest to system operators for deciding if either parts have to be replaced or only have to be recelebrated. Because measured data on system deviations
often reflects complex intertwining causes and \textit{closed-loop corrections}, identifying root causes becomes a matter of inference. The proposed two-step approach 
first builds a generative model (based on control software and expert insights) that simulates observations from hypothesized error causes. This simulator is then recast 
as a probabilistic program, allowing Bayesian inference to estimate latent causes (with confidence intervals) from actual measurements.

@@

% Full Miniprob grammar here.
